---
title: "mapmaker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mapmaker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

The package mapmaker includes a single function ('mapmaker') designed to draw maps based on GSHHS-NOAA coastline data.

### Required Data and Data Example

<br>

GSHHS-NOAA data can be downloaded at https://gnome.orr.noaa.gov/goods/tools/GSHHS/coast_subset}. The data download as .bna files (Atlas Boundary Files) with shoreline coordinates defined as hierarchically arranged closed polygons.

<br>

Once downloaded from the NOAA website, the .bna file needs to be uploaded into a data.frame that can be used by the mapmaker function. This can be done as follows.

<br>

object <- read.delim('filename.bna', sep=',', header=TRUE)

<br>

An example data (Bahamas) is included in the package. It is a data.frame created from a GSHHS-NOAA .bna file.

### Package Installation

The package mapmaker can be installed from github using devtools.

<br>

devtools::install_github(`"`mjkowalewski/mapmaker`"`, build_vignettes = TRUE)

<br>

Once installed, the library can be uploaded.

```{r setup}
library(mapmaker)
```

### mapmaker function

The function *mapmaker* is designed to either draw a map locally in the R/R Studio plot window (default) or export an external pdf file (when argument pdf = TRUE). The default setting is pdf = FALSE because it is advisable to preview a map before producing the final output. The plot is for preview purpose only because the map aspect ratio will not be correct, but will reflect the plot window dimensions and the longitudinal and latitudinal ranges of the map. When the pdf = TRUE, the exported pdf file will have the correct map aspect ratio. 

To explore the function, the example dataset "bahamas" will be used here. 

```{r example data}
str(bahamas)
head(bahamas)
```

Here is an example of a map produced by the function for the dataset "bahamas" using the function's default settings.The pdf version of this figure offers much higher quality and resolution than the figure in the html version rendered here.  

```{r creating a map, echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(bahamas)
```

The map can be customized by changing colors and modifying the placement and attributes of the scale bar and the north arrow. Both, the scale bar and the north arrow can be suppressed. See help menu ?mapmaker for details about all currently available arguments.

<br>

Here is an example of a customized map for the Bahamas dataset.

```{r a map with arguments, echo=TRUE, fig.dim = c(6, 6), fig.align='center'}
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5,
         arr.cex = 1.1, arr.coord = c(-73, 26.5, 27.5),
         sea.col='turquoise', land.col='yellow', 
         coast.col = 'gray50')
```


Finally, additional elements can be added to the map such as towns or sampling sites using 'rscript={}' argument.The package provides example data sets ('bahamas.towns' and 'bahamas.sites'), which are used in the example below.

```{r map customizing, echo=TRUE, fig.dim = c(6, 6), fig.align='center'}
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5, 
         arr.cex = 0.8, arr.coord = c(-73, 26.5, 27.5), 
         rscript={
           mtext(side=3, line=0.1, "my map")
           points(bahamas.towns[,2:3], pch=24, cex=0.7,
                  col='black', bg='white')
           text(bahamas.towns[,2:3], cex=0.5, 
                col='gray30', bahamas.towns[,1],
                pos = 2, offset=0.15)
           points(bahamas.sites[,2:3], pch=21, cex=1.5,
                  col='black', bg='white')
           text(bahamas.sites[,2:3], cex=0.5, col='firebrick',
                bahamas.sites[,1])
})
```
<br>

Once the map is finalized, the map can be exported by setting the pdf = TRUE.

<br>

**Notes and Comments**

  * The resolution of a map drawn by mapmaker will depend on the resolution chosen when downloading the .bna file from NOAA website.
  
  * For maps crossing the dateline longitude west values are > 180.
  
  * The map is always produced in the northward orientation.
  
  * The function computes map aspect ratio based on the mid-point latitude of a given map. The aspect ratio only applies to pdf-formatted figures produced by mapmaker function when argument pdf = TRUE.
  
  * The scaling is not applied When the map is previewed in R plot window (pdf = FALSE).

<br>

**Comments? Question?** Michal Kowalewski (University of Florida) kowalewski@ufl.edu
