---
title: "mapmaker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mapmaker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

The package mapmaker includes a single function ('mapmaker') designed to draw maps based on GSHHS-NOAA coastline data.

### Required Data and Data Example

<br>

GSHHS-NOAA data can be downloaded at https://gnome.orr.noaa.gov/goods/tools/GSHHS/coast_subset}. The data download as .bna files (Atlas Boundary Files) with shoreline coordinates defined as hierarchically arranged closed polygons.

<br>

Once downloaded from the NOAA website, the .bna file needs to be uploaded into a data.frame that can be used by the mapmaker function. This can be done as follows.

<br>

object <- read.delim(`'filename.bna'`, sep=`','`, header=TRUE)

<br>

Three example data sets (bahamas, adriatic, bearing.straits) are included in the package. All three are dataframes created from GSHHS-NOAA .bna files.

### Package Installation

The package mapmaker can be installed from github using devtools.

<br>

devtools::install_github(`"mjkowalewski/mapmaker"`, build_vignettes = TRUE)

<br>

Once installed, upload the library.

```{r setup}
library(mapmaker)
```

### mapmaker function

The function *mapmaker* is designed to either draw a map locally in the R/R Studio plot window (pdf = FALSE) or export an external pdf file (pdf = TRUE). The default setting is pdf = FALSE because, in most cases, a map needs to be previewed and adjusted before generating the final pdf output. The plot is for preview purposes and the map aspect ratio will not be correct, but reflects the plot window dimensions and the longitudinal and latitudinal ranges of the map. The exported pdf file will have dimensions with map aspect ratio based on the mid-point latitude of the map. In its default form, the map will include North Arrow (plotted in the upper right corner of the map) and the scale bar (plotted in the top center part of the map). Both, the arrow and the scale bar can be suppressed or modified using arrow and scale arguments included in the function. Also, custom dimensions can be defined to customize figure dimensions, although using this argument will usually result in producing a map with incorrect aspect ratios.

The example datasets included in the package are dataframes with three columns. All have the same format exemplified here using data 'bahamas'. 

```{r example data}
str(bahamas)
head(bahamas)
```

Below is an example of a map drawn for the dataset 'adriatic' using default version of the function 'mapmaker'. Note that the resolution of the html version of the map is relatively low. The pdf version of this figure generated in R should offer a much higher quality and resolution.

```{r creating a map (adriatic), echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(adriatic)
```

A map also can be drawn using grayscale by setting the argument grayscale = TRUE (default: grayscale = FALSE) or by customizing color arguments using grayscale color definitions.

```{r creating a map (bahamas), echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(bahamas, grayscale = TRUE)
```

Note that if the map includes the dateline, the longitude West values are positive > 180 rather than negative (-180 - 0), as illustrated here for Bearing Strait.

```{r creating a map (bearing strait), echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(bearing.straits)
```

The map can also be customized by changing colors of land, sea, and coastline,  modifying the arguments for the scale bar and the north arrow (see help menu (`?mapmaker`) for details about all currently available arguments). Here is an example, using 'bahamas' dataset.

<br>

```{r a map with arguments, echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5,
         arr.cex = 1.1, arr.coord = c(-73, 26.5, 27.5),
         sea.col='turquoise', land.col='yellow', 
         coast.col = 'gray50')
```

Additional elements can be added to the map such as towns or sampling sites using 'rscript={}' argument.The package provides example data sets ('bahamas.towns' and 'bahamas.sites'), which are used in the example below.

```{r map customizing, echo=TRUE, fig.dim = c(5, 5), fig.align='center'}
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5, 
         arr.cex = 0.8, arr.coord = c(-73, 26.5, 27.5), 
         grayscale = TRUE,
         rscript={
           mtext(side=3, line=0.1, "my map")
           points(bahamas.towns[,2:3], pch=24, cex=0.7,
                  col='black', bg='white')
           text(bahamas.towns[,2:3], cex=0.5, 
                col='gray30', bahamas.towns[,1],
                pos = 2, offset=0.15)
           points(bahamas.sites[,2:3], pch=21, cex=1.5,
                  col='black', bg='white')
           text(bahamas.sites[,2:3], cex=0.5, col='firebrick',
                bahamas.sites[,1])
})
```
<br>

<br>

Once the map is finalized, the map can be exported by setting the argument pdf = TRUE.

<br>

**Notes and Comments**

  * The resolution of a map drawn by mapmaker will depend on the resolution chosen when downloading the .bna file from NOAA website.
  
  * For maps crossing the dateline longitude west values are > 180.
  
  * The map is always produced in the northward orientation.
  
  * The function computes map aspect ratio based on the mid-point latitude of a given map. The aspect ratio only applies to pdf-formatted figures produced by mapmaker function when argument pdf = TRUE.
  
  * The scaling is not applied when the map is previewed in R plot window (pdf = FALSE).

<br>

**Comments? Question?** Michal Kowalewski (University of Florida) kowalewski@ufl.edu
