---
title: "mapmaker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mapmaker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

The package mapmaker is designed to draw maps based on GSHHS-NOAA coastline data.

### Required Data and Data Example

<br>

GSHHS-NOAA data can be downloaded at https://gnome.orr.noaa.gov/goods/tools/GSHHS/coast_subset}. The data download as .bna files (Atlas Boundary Files) with shoreline coordinates defined as hierarchically arranged closed polygons.

<br>

Once downloaded from the NOAA website, the .bna file needs to be uploaded into a data.frame that can be used by the mapmaker function. This can be done as follows:

<br>

object <- read.delim(`'filename.bna'`, sep=`','`, header=TRUE)

<br>

Three example data sets (bahamas, adriatic, bearing.straits) are included in the package. All three are dataframes created from GSHHS-NOAA .bna files.

### Package Installation

The package mapmaker can be installed from github using devtools.

<br>

devtools::install_github(`"mjkowalewski/mapmaker"`, build_vignettes = TRUE)

<br>

Once installed, upload the library.

```{r setup}
library(mapmaker)
```

### mapratio function

The function computes the width and height of the map with the corrected aspect ratio. The dimensions are corrected for differences in ranges of latitudes and longitudes and for longitude/latitude ratio at a given latitude using mid-point latitude of the map. Note that longitude/latitude is 1 at the equator and <1 at higher latitudes.

<br>

The function returns a vector with two numerical values, representing width and height of map in inches. In its default form, width = 3 inches and height value is computed by the function. If height is defined using argument map.height, the width is computed by the function instead. For example:

```{r mapratio output examples}
mapratio(bahamas) # dimensions for a map width of 3 inches (default)
mapratio(bahamas, map.width = 4) # dimensions for map width of 4 inches
mapratio(bahamas, map.height = 2.5) # dimensions for map with height of 2.5 inches
```

### mapmaker function 

The function *mapmaker* draws a single plot (map). In its default form, the map will include North Arrow (plotted in the upper right corner of the map) and the scale bar (plotted in the top center part of the map). Both, the arrow and the scale bar can be suppressed or modified using arrow and scale arguments.

The object coords (required) should be a data frame with three columns derived from a .bna file. All example datasets included in the package conform to this format, illustrated here using data 'bahamas'. The first four rows define map boundaries and are used by *mapratio* to compute map dimensions and *mapmaker* to draw a map.

```{r example data}
str(bahamas)
head(bahamas)
```

Below are examples of maps drawn using example datasets. Note that mapratio function is implemented first to obtain map dimensions, which are then used to controlmap dimensions (e.g., using argument *pin* in *par* function). The resolution of the example maps included in this vignette is necessarily low, but figures with much higher (publication quality) maps can be produced in R.

```{r creating a map (adriatic), echo=TRUE, fig.dim = c(5, 5), fig.align='center', dpi = 150}
adriatic.dim <- mapratio(adriatic)
par(pin = adriatic.dim)
mapmaker(adriatic)
```

A map also can be drawn using grayscale by setting the argument grayscale = TRUE (default: grayscale = FALSE) or by customizing color arguments using grayscale color definitions.

```{r creating a map (bahamas), echo=TRUE, fig.dim = c(5, 5), fig.align='center', dpi = 150}
bahamas.dim <- mapratio(bahamas)
par(pin = bahamas.dim)
mapmaker(bahamas, grayscale = TRUE)
```

Note that if the map includes the dateline, the longitude West values are positive > 180 rather than negative (-180 - 0), as illustrated here for the Bearing Strait map.

```{r creating a map (bearing strait), echo=TRUE, fig.dim = c(5, 5), fig.align='center', dpi=150}
bstraits.dim <- mapratio(bearing.straits)
par(pin = bstraits.dim)
mapmaker(bearing.straits)
```

The map can also be customized by changing colors of land, sea, and coastline,  modifying the arguments for the scale bar and the north arrow (see help menu (`?mapmaker`) for details about all currently available arguments). Here is an example, using 'bahamas' dataset.

<br>

```{r a map with arguments, echo=TRUE, fig.dim = c(5, 5), fig.align='center', dpi = 150}
bahamas.dim <- mapratio(bahamas)
par(pin = bahamas.dim)
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5,
         arr.cex = 1.1, arr.coord = c(-73, 26.5, 27.5),
         sea.col='turquoise', land.col='yellow', 
         coast.col = 'gray50')
```

Additional elements can be added to the map such as towns or sampling sites using 'rscript={}' argument.The package provides example data sets ('bahamas.towns' and 'bahamas.sites'), which are used in the example below.

```{r map customizing, echo=TRUE, fig.dim = c(5, 5), fig.align='center', dpi = 150}
bahamas.dim <- mapratio(bahamas)
par(pin = bahamas.dim)
mapmaker(bahamas, scale.width = 50, scale.lwd = 1.5,
         scale.lat=27, scale.long=-76, arr.lwd = 1.5, 
         arr.cex = 0.8, arr.coord = c(-73, 26.5, 27.5), 
         grayscale = TRUE,
         rscript={
           mtext(side=3, line=0.1, "my map")
           points(bahamas.towns[,2:3], pch=24, cex=0.7,
                  col='black', bg='white')
           text(bahamas.towns[,2:3], cex=0.5, 
                col='gray30', bahamas.towns[,1],
                pos = 2, offset=0.15)
           points(bahamas.sites[,2:3], pch=21, cex=1.5,
                  col='black', bg='white')
           text(bahamas.sites[,2:3], cex=0.5, col='firebrick',
                bahamas.sites[,1])
})
```

<br>

**Notes and Comments**

  * The resolution of a map drawn by mapmaker will depend on the resolution chosen when downloading the .bna file from NOAA website.
  
  * For maps crossing the dateline longitude west values are > 180.
  
  * The map is always produced in the northward orientation.
  
  * The *mapratio* function computes map aspect ratio based on the mid-point latitude of a given map.
  
<br>

**Comments? Question?** Michal Kowalewski (University of Florida) kowalewski@ufl.edu
